Scaling up a GF grammar and preparing it for cool stuff for Swedish

### 2011-12-01 doc/report/report.tex
report and pictures

\documentclass{report} 
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[margin=1.5in]{geometry}
\usepackage{lingmacros}
\usepackage{verbatim}
\usepackage{subfig}
% exempel: 2) Inte vet jag
%              not know I
%             I don't know

@@

\subsection{Swedish}
Swedish\cite{SAG:intro} is a North-Germanic language,
closely related to Norweigan and Danish. The languages share most of their
grammatical structures and are mutual intelligibility. Swedish is also 
an official language in Finland and spoken by approximately 10 million people.
Swedish syntax are often similair to English, but have some characteristics:
\subsubsection*{Verb second}
Swedish is a verb-second language: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but any syntactic category can be fronted\cite{H&H:1027}.
Topicalisation is very common, especially for temporal and spatial adverbial phrases.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, although the fronted part of sentence \ref{ex:swedish-ovs} and \ref{ex:swedish-avso} are being emphasized.
\enumsentence{
\shortex{4}
{Du & ser & mig & inte.}
{You & see & me & not.}
{`You  don't see me'.}} \label{ex:swedish-svo}
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Mig  & ser &  du & inte.}
{Me & see & you & not.}}\label{ex:swedish-ovs} 
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Inte & ser& du &mig.}
{Not&  see& you& me.}} \label{ex:swedish-avso}

@@

\subsubsection*{Impersonal constructions}
Constructions with 'det är/var' ('it is/was') is very common\cite{H&H309d}:
\enumsentence{
\shortex{6}{Det & var & roligt & att & höra.}
{It &was & nice &to &hear.}
{`I'm glad to hear that.'}}
'Det' (\emph{it}) is also used as formal subject %for intransitive verbs,
with the real subject put in the position of an object.
\enumsentence{
\shortex{6}{Det &står &en &älg &på &fältet.}
{ It & stands &a & moose &on &the field.}
{`There is a moose on the field.'}}

\subsubsection*{Reflexive pronouns}
\label{swe:refl}
The Scandinavian language have reflexive pronouns\cite{H&H310}
and special reflexive forms of the possessive pronouns for the 3rd
person\cite{H&H319}. Those are distinct from the normal 3rd person forms.
\enumsentence{
\begin{tabular}{ll}
Han slog \textbf{sig}. & Han såg \textbf{sitt} barn.\\
\vspace{3mm}
\emph{He hit him self.} & \emph{He saw his (own) child.}\\
Han slog \textbf{honom}. & Han såg \textbf{hans} barn.\\
\emph{He hit him (another person).}&\emph{He saw his (another person's) child.}
\end{tabular}}
The 1st and 2nd persons use the normal personal pronoun in object form.

@@

\subsection{Grammatical Framework}
The main component of the the project is the Grammatical Framework\ref{gfbok} (GF). It is
a functional grammar formalism based on Martin-Löf type theory.
It has been developed for about 20? years and are inspired by programming langauges
like Haskell\footnote{}, ML\footnote{}, Prolog\footnote{} and Agda\footnote{}. Grammars written in GF can be used together with languages 
like Java, Java script and Haskell and works for both parsing and generation.

@@

A GF grammar usually covers a very restricted set of language where few ambiguities
occur and much care can be taken to preserving the semantics of the
different languages.
By using a controlled language like this,  good and reliable translation can be
ensured. Some examples of projects using GF are the math .. \cite{olga}, the
Online Syllogism Solver\cite{malin!}.  The framework is also used in the
european project MOLTO for online translation between upto 15 language.\\

@@

As an open source-project, GF is constantely being developed and improved. New
languges are added, the compiler is being improved, ways of använad det på ett
lätt sätt are added and the possiblities to use GF in different environments
increased. There are  research about how to make better use of the dependent
types, for solving syllogism or generating natural language via Montague
semantics.
Further, experiments on using GF for free parsing has been conducted (Angelov,2010?).
This is also a future goal for our project and one of the delmål has been to
model an important part of Swedish in GF. The starting point was the resource
grammar and the result should still have a strong connection to this.

@@

\subsection{Talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being an freely avaible, manually annotade, large scale treebank,
Talbanken\cite{talbanken} was perfect for this ändamål.
It was created in the 1970s at Lund University. In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school (?) students, and the last one, section P,
consists of professionally written Swedish gathered from newspapers and ... .\\
Talbanken was also used to train the Swedish version of the Maltparser\cite{mal}
and was därmed reutgiven in a more modern version\cite{nivre}. This work
was done by Nivre and ... in 2005. They released it in Malt and Tiger XML-formats,
where the trees had been made deeper and smt more. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Related work}
Computational ling is active area for Swedish, there are many other Swedish parser,
but grammars?
Many of the parsers not ruled based, the most known one Malt ... .
However has been two, about CLE and what happend to it. Logic and translation
to Swedish, The spoken lang. translator.
Other Sw. parsers and grammars, evaluation of them? 
Other grammar frameworks.
% related and references work: CLE, Krasimirs book, CassSwe, GF book,SAG, gunlög,
% the multilingual treebank, Homse&Hinchcliff, Aarne i LILT

\chapter{Proceedings}
%\section{Proceedings med bättre ord}  
The purpose has been to prepare the grammar for large scale parsing.
Develop method used for English to fit Swedish, pynta the sw grammar with
language specific constructions and create methods for lexicon.
Earlier have done tool, but the lexicon was not big enough to be able
to fix it in this way. Tool may instead be used for complementing.

\section{Saldo}
\label{sec:prog.saldo}
In saldo there is also semantic info, so
We need the numbers from saldo if there are to lemmas with the same
identifier. sluta\_V and sluta2\_V (slöt). However gf does not want more than one table
if the forms are identical, saldo's morpho neither, but there are five in big saldo. 
We are just interested in morphology, so far. Dont want too big lexicon.

Saldo has been imported before (Krasimir, 2009?), but the code needed updates
to not miss many important words.

The basic algorith is the same. To understand this need to look at differences.
Example of gf and saldo table.
Explanation of the algorithm: map categories, try each paradigm, compare the table of forms
created. If several forms in saldo, gf should pick on of them. Saldo has forms for
compounds, gf does not. A grammar is written, the correct ones saved, the others 
retried. Try all paradigms which can be formed from saldo. 
There are a number of differences between and saldo Till skillnad från GF,
saldo does not contain forms that aren't used gf generates all forms. A form
needed to generate the gf table may not be given in saldo such as singular
forms for 'byxor'. Also contains idioms, 'hålla huvudet kallt' and has
different grammatical analyse, a lot of pronouns. Example.

@@

\section{The grammar}
Although we can't expect to get a full coverage, we would like a tillfredsställande
Explain what's in the resources, constructions present in most languages. 
Been working on a grammar fragment based on Talbanken,
like cool people like Montague. May also be interesting for other purposes.
Swedish shares with Nor and Dan. Extra module, covering things like sw is a
v2-lang, focusing parts of sentences.

@@

For verbs, but also others there is an important distinction for valency.
Either directly to VP if V, or needs to be given arguments. Can then add adverbs.
Maybe talk about ställa.
Particles and prepositions, can't be seen on the trees.
tycka om /tycka om -> om vad tycker du inte. fast bra

@@

%% Clauses
\subsubsection{Impersonal constructions}%Presentering, , Topicalisering
\label{sec:Formal}
Formal subjects \cite{SAG-19,H&H-309d} is often used in Swedish.
\enumsentence{\shortex{6}
{Det & sitter & en & fågel & på & taket}
{It & sits & a & bird & on & the roof}
{``There is a bird sitting on the roof''}}
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.
Transitive verbs may not be used like this
\enumsentence{
*Det äter en fågel på taket\\
 ''There is a bird eating on the roof"}

@@

\subsubsection{Reflexive pronouns}
%varandra and sin/sitt/sig.
An important piece in a Swedish grammar is the reflexive pronouns and
%both \emph{sin} and , eg. \emph{sin}.
the possessive pronouns as described in section \ref{swe:refl}.
Those make up noun phrases wich are not allowed in subjects, 
and we want to be able to parse to following sentences:
\enumsentence{Han såg sina barns skor.\\ \label{ex:ReflGen}
He saw his children's shoes.}
\enumsentence{Sina vantar hade han glömt på tåget.\\
His gloves, he had forgotten on the train.}
\enumsentence{Han ger sina pengar till sina syskon.\\
He gives his money to his children.}
\enumsentence{Han är längre än sin kompis.\\
He is taller than his friend.} \label{ex:ReflAP}
\vspace{-2mm}
\enumsentence{Han är här oftare än sin kompis.\\
He is here more often than his friend.} \label{ex:ReflAdv}
\vspace{-2mm}
\enumsentence{Hon tyckte om skolan och alla sina elever.\\
She liked the school and all her students.}\label{ex:ReflConj}
\enumsentence{Han såg sina få böcker och sin penna.\\
He saw his few books and his pencil.}\label{ex:ReflConj2}
\vspace{-2mm}
\enumsentence{Hon ber alla sina kompisar att gå\\
She asks all her friends to leave}\label{ex:ReflPredet}
\enumsentence{Jag vill att han rakar sig.\\
I want him to shave.}
but not
\enumsentence{*Sina vantar var kvar på tåget.\\
Her gloves were left on the train.}\label{ex:ReflSubj}
\enumsentence{*Jag ger sina pengar till sina syskon.\\
I give his money to his siblings. }
\enumsentence{*Han vill att jag rakar sig.\\
He wants me to shave himself.}

@@

here or in robust: SupCl, det as VP or VV or S for S (remember Foc)
\subsubsection{Relative clauses}
The resource grammar already provided a number of ways to express relative and embedded clauses, such as
\enumsentence{Frågan är vilken färg den har\\
                   Pojken, som är blyg, tystnar\\
                   Han såg kunden som tyckte om sallad\\
                   Jag tänkte på huset i vilket hon bodde} % better example!!

@@

A program for translating trees to GF syntax has been done before (Angelov,).
It translated trees from the English Penn Treebank. By starting this progam, we now have
a translation that works for the quite differently annotated Talbanken. The differences between
Swedish and English syntax also meant that many rules had to be changed.\\

@@

\end{document}

>>>>

\documentclass{report} 
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[margin=1.5in]{geometry}
\usepackage{lingmacros}
\usepackage{verbatim}
\usepackage{subfig}
\usepackage{wrapfig}
% exempel: 2) Inte vet jag
%              not know I
%             I don't know

@@

\subsection{Swedish}
Swedish\cite[Inl. \textsection 3]{SAG}is a North-Germanic language,
closely related to Norweigan and Danish. The languages share most of their
grammatical structures and are mutual intelligibility. Swedish is also 
an official language in Finland and spoken by approximately 10 million people.
Swedish syntax are often similair to English, but have some characteristics:
\subsubsection*{Verb second}
Swedish is a verb-second language\cite[p. 116]{gunlog}: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but any syntactic category can be fronted\cite[\textsection 1027]{H&H}.
Topicalisation is very common, especially for temporal and spatial adverbial phrases.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, although the fronted part of sentence \ref{ex:swedish-ovs} and \ref{ex:swedish-avso} are being emphasized.
\enumsentence{
\shortex{4}
{Du & ser & mig & inte.}
{You & see & me & not.}
{`You  don't see me'.}} \label{ex:swedish-svo}
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Mig  & ser &  du & inte.}
{Me & see & you & not.}}\label{ex:swedish-ovs} 
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Inte & ser& du &mig.}
{Not&  see& you& me.}} \label{ex:swedish-avso}

@@

\subsubsection*{Impersonal constructions}
Constructions with 'det är/var' ('it is/was') is very common\cite[\textsection 309d]{H&H}:
\enumsentence{
\shortex{6}{Det & var & roligt & att & höra.}
{It &was & nice &to &hear.}
{`I'm glad to hear that.'}}
'Det' (\emph{it}) is also used as formal subject %for intransitive verbs,
with the real subject put in the position of an object.
\enumsentence{
\shortex{6}{Det &står &en &älg &på &fältet.}
{ It & stands &a & moose &on &the field.}
{`There is a moose on the field.'}}

\subsubsection*{Reflexive pronouns}
\label{swe:refl}
The Scandinavian language have special reflexive pronouns
and possessive pronouns for the 3rd
person\cite[\textsection 310 \& 319]{H&H}. Those are distinct from the normal 3rd person forms.
\enumsentence{
\begin{tabular}{ll}
Han slog \textbf{sig}. & Han såg \textbf{sitt} barn.\\
\vspace{3mm}
\emph{He hit him self.} & \emph{He saw his (own) child.}\\
Han slog \textbf{honom}. & Han såg \textbf{hans} barn.\\
\emph{He hit him (another person).}&\emph{He saw his (another person's) child.}
\end{tabular}}
The 1st and 2nd persons use the normal personal pronoun in object form.

@@

\subsection{Grammatical Framework}
The main component of the the project is the Grammatical Framework\cite{gfbok} (GF). It is
a functional grammar formalism based on Martin-Löf type theory.
It has been developed for about 20? years and are inspired by programming langauges
like Haskell\cite{haskell},
ML\cite{ml}, Prolog\cite{bratko} and
Agda\cite{agda}.
Grammars written in GF can be used together with languages 
like Java, Java script and Haskell and works for both parsing and generation.

@@

A GF grammar usually covers a very restricted set of language where few ambiguities
occur and much care can be taken to preserving the semantics of the
different languages.
By using a controlled language like this,  good and reliable translation can be
ensured. Some examples of projects using GF are WebAlt\cite{webalt} and something else.
%, the Online Syllogism Solver\cite{malin!}. 
The framework is also used in the
european project MOLTO\cite{molto} for online translation between upto 15 language.\\

@@

As an open source-project, GF is constantely being developed and improved. New
languges are added, the compiler is being improved, ways of använad det på ett
lätt sätt are added and the possiblities to use GF in different environments
increased. There are  research about how to make better use of the dependent
types, for solving syllogism or generating natural language via Montague
semantics.
Further, experiments on using GF for free parsing has been conducted \cite[\textsection 4.1]{gfMech} 
This is also a future goal for our project and one of the delmål has been to
model an important part of Swedish in GF. The starting point was the resource
grammar and the result should still have a strong connection to this.

@@

\subsection{Talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being an freely avaible, manually annotade, large scale treebank,
Talbanken\cite{talbanken} was perfect for this ändamål.
It was created in the 1970s at Lund University. In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school students, and the last one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser\cite{malt}
and was thus redistributed in a more modern version\cite{talbanken05} This work
was done by Nivre, Hall and Nilsson in 2005. They released it in Malt and Tiger XML-formats,
where the trees had been made deeper and more detailed. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Related work}
Computational ling is active area for Swedish, there are many other Swedish parser,
but grammars?
Many of the parsers not ruled based, the most known one Malt\cite{malt} ... .
However has been two, about CLE\cite{cle} and what happend to it. Logic and translation
to Swedish, The spoken lang. translator\cite{cle2}. Wiren \cite{wiren}, cassSwe \cite{casswe}
Other Sw. parsers and grammars, evaluation of them? 
Other grammar frameworks.
% related and references work: CLE, Krasimirs book, CassSwe, GF book,SAG, gunlög,
% the multilingual treebank, Homse&Hinchcliff, Aarne i LILT

\chapter{Proceedings}
%\section{Proceedings med bättre ord}  
The purpose of this project has been to prepare the grammar for free parsing.
This has meant to develop earlier techniques to fit for Swedish, create methods
for achieving and keeping a extensive lexicon and to adapt the existing
resource grammar to fit for more language specific constrictions.
The methodology and results will be presented in the following section.

\section{Saldo}
\label{sec:prog.saldo}
Saldo is compatible with GF, which makes it easy to transform the morphological
lexicon into our format. The lexicon has been imported earlier (Krasimir,2009?)
but the code needed updates to not miss many important words.
Experiments have been made to use a interactive tool for lexical aquisation,
but the dictionary was still much to small. The aquisation tool may still
be used for complenting the lexicon.

The basic algorithm for porting Saldo is the same as Angelov's. It makes use of
the smart paradigms in GF : explanation.
%Find the paper!
%To understand it, we first need to look at differences between Saldo and a GF lexicon.
The program will try to feed as few forms as possible to the GF smart paradigm.
It starts from the easiest case, compare the results and either saves the code or 
makes another try using more word forms.
Figure \ref{pic:TabVax} shows a the information given by Saldo and by GF respectively.\\
\begin{figure}[h]
  \begin{center}
\subfloat[Saldo]{\label{pic:saldoVax}\includegraphics[height=90mm]{saldoVax.png}}
\hspace{5mm}
\subfloat[GF]{\label{pic:gfVax}\includegraphics[height=60mm]{gfVax.png}}
\caption{Inflection table for the verb ``växa" (grow)}
\label{pic:TabVax}
  \end{center}
\end{figure}
We can see that there are more saldo forms and some are only in GF.

Explanation of the algorithm: map categories, try each paradigm, compare the table of forms
created. If several forms in saldo, gf should pick on of them. Saldo has forms for
compounds, gf does not. A grammar is written, the correct ones saved, the others 
retried. Try all paradigms which can be formed from saldo. 
There are a number of differences between and saldo Till skillnad från GF,
saldo does not contain forms that aren't used gf generates all forms. A form
needed to generate the gf table may not be given in saldo such as singular
forms for 'byxor'. Also contains idioms, 'hålla huvudet kallt' and has
different grammatical analyse, a lot of pronouns. Example.

In Saldo there is also semantic info, so
We need the numbers from saldo if there are to lemmas with the same
identifier. växa (bli större), växa (vara en planta) vs. sluta\_V and sluta2\_V
(slöt). However gf does not want more than one table
if the forms are identical, saldo's morpho neither, but there are five in big saldo. 
We are just interested in morphology, so far. Dont want too big lexicon.

@@

\subsection{Results}
How many words, how many Talbanken words are not here now? Why?

\section{The grammar}
\label{sec:prog.grammar}
Although we can't expect to get a full coverage, we would like a tillfredsställande
Explain what's in the resources, constructions present in most languages. 
Been working on a grammar fragment based on Talbanken,
like cool people like Montague. May also be interesting for other purposes.
Swedish shares with Nor and Dan. Extra module, covering things like sw is a
v2-lang, focusing parts of sentences.

@@

For verbs, but also others there is an important distinction for valency.
Either directly to VP if V, or needs to be given arguments. Can then add adverbs.
Maybe talk about ställa.
Particles and prepositions, can't be seen on the trees.
tycka om /tycka om -> om vad tycker du inte. fast bra
about slashes\cite{gazdar}

@@

%% Clauses
\subsubsection{Impersonal constructions}%Presentering, , Topicalisering
\label{sec:Formal}
Formal subjects \cite[19]{SAG} is often used in Swedish.
\enumsentence{\shortex{6}
{Det & sitter & en & fågel & på & taket}
{It & sits & a & bird & on & the roof}
{``There is a bird sitting on the roof''}}
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.
Transitive verbs may not be used like this
\enumsentence{
*Det äter en fågel på taket\\
 ''There is a bird eating on the roof"}

@@

here or in robust: SupCl, det as VP or VV or S for S (remember Foc)
\subsubsection{Reflexive pronouns}
%varandra and sin/sitt/sig.
An important piece in a Swedish grammar is the reflexive pronouns and
%both \emph{sin} and , eg. \emph{sin}.
the possessive pronouns as described in section \ref{swe:refl}.
Those make up noun phrases wich are not allowed in subjects, 
and we want to be able to parse to following sentences:
\enumsentence{Han såg sina barns skor.\\ \label{ex:ReflGen}
He saw his children's shoes.}
\enumsentence{Sina vantar hade han glömt på tåget.\\
His gloves, he had forgotten on the train.}
\enumsentence{Han ger sina pengar till sina syskon.\\
He gives his money to his children.}
\enumsentence{Han är längre än sin kompis.\\
He is taller than his friend.} \label{ex:ReflAP}
\vspace{-2mm}
\enumsentence{Han är här oftare än sin kompis.\\
He is here more often than his friend.} \label{ex:ReflAdv}
\vspace{-2mm}
\enumsentence{Hon tyckte om skolan och alla sina elever.\\
She liked the school and all her students.}\label{ex:ReflConj}
\enumsentence{Han såg sina få böcker och sin penna.\\
He saw his few books and his pencil.}\label{ex:ReflConj2}
\vspace{-2mm}
\enumsentence{Hon ber alla sina kompisar att gå\\
She asks all her friends to leave}\label{ex:ReflPredet}
\enumsentence{Jag vill att han rakar sig.\\
I want him to shave.}
but not
\enumsentence{*Sina vantar var kvar på tåget.\\
Her gloves were left on the train.}\label{ex:ReflSubj}
\enumsentence{*Jag ger sina pengar till sina syskon.\\
I give his money to his siblings. }
\enumsentence{*Han vill att jag rakar sig.\\
He wants me to shave himself.}

@@

\subsubsection{Relative clauses}
The resource grammar already provided a number of ways to express relative and embedded clauses, such as
\enumsentence{Frågan är vilken färg den har\\
                   Pojken, som är blyg, tystnar\\
                   Han såg kunden som tyckte om sallad\\
                   Jag tänkte på huset i vilket hon bodde} % better example!!

@@

A translation of the English Penn Treebank to the GF syntax has previosly been done \cite{gfpenn}.
By starting this progam, we now have
a translation that works for the quite differently annotated Talbanken. The differences between
Swedish and English syntax also meant that many rules had to be changed.\\

@@

\bibliographystyle{ieeetr}
\bibliography{masterbib}
\end{document}
